Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions

### 2022-11-15 SIMPaper/SIMPaper.Rmd
addedStatTemplate

Now, let $b$ be the (user-defined) size of the *base series* and let $B$ be the sum of all follow-up times for the individuals in the study. If we sample the *base series* uniformly across the study base, then the hazard function of the sampling process is equal to $b/B$. Therefore, we have the following equality \footnote{We are abusing notation here, conflating hazards with probabilities. For a rigorous treatment, see Saarela \& Hanley (2015) section 3 \cite{saarela2015} .}: 
\begin{align}\label{eqn:main}
\frac{P\left(Y_i=1 \mid X_i, T_i\right)}{P\left(Y_i = 0 \mid X_i, T_i\right)} = \frac{h\left(T_i \mid X_i\right)}{b/B}.
\end{align} 
The odds of a person-moment being a part of the *case series* is the ratio of the hazard $h(T_i \mid X_i)$ and the uniform rate $b/B$. Using \eqref{eqn:main}, we can see how the log-hazard function can be estimated from the log-odds arising from case-base sampling: \begin{align}\label{eqn:offset}
\log \left( h\left(t \mid X_i\right)\right) = \log \left(\frac{P\left(Y_i = 1 \mid X_i, t\right)}{P\left(Y_i = 0 \mid X_i, t\right)}\right) + \log\left(\frac{b}{B}\right).
\end{align}

To estimate the correct hazard function, we adjusting for the bias introduced when sampling a fraction of the study base $B$ by adding the offset term $\log\left(\frac{b}{B} \right)$ as in \eqref{eqn:offset}. Next, we propose using neural networks to model the odds.

@@

After case-base sampling, we pass all features, including time, into any user-defined feed-forward component, to which an offset term is added, then passed through a sigmoid activation function (Figure \ref{fig:NNarch}). We use the sigmoid function as its inverse is the odds, which we can use to calculate the hazard. The general form for the neural network using CBNN is:

@@

where $T$ is a random variable representing the event time, $X$ is the random variable for a covariate profile, $f_{\theta}(X, T)$ represents any feed-forward neural network architecture, $\log\left(\frac{B}{b}\right)$ is bias term set by case-base sampling, $\theta$ is the set of parameters learned by the neural network and $\sigmoid(x)=\frac{1}{1+e^{-x}}$. By approximating a higher-order polynomial of time using a neural network, the baseline hazard specification is now data-driven, where user-defined hyperparameters such as regularization, number of layers and nodes control the flexibility of the fukk hazard function. We provide a detailed description of the choices we made in the next sub-section.

@@

We use two metrics to assess the performance of the different methods of interest: 1) (IPA) [@kattan2018index] and 2) inverse probability censoring weights-adjusted concordance index ($C_{IPCW}$) [@uno2011], which we define below. Both these time-dependent metrics provide transparency as to when in follow-up time each model may perform better than the others.

@@

```{r megaPlot, echo = FALSE , fig.align = 'center', out.width = "100%", fig.cap = "Summarizes the simple simulation (A, B), complex simulation (C, D), SUPPORT case study (E, F) and METABRIC case study (G, H) results. The first row shows the IPA score for each model in each study over follow-up time. Negative values mean our model performs worse than the null model and positive values mean the model performs better. The second row demonstrates the $C_{IPCW}$ score for each model in each study over follow-up time. A score of 1 is the maximum performance for either metric. Each model-specific metric in each study shows a 95-percent confidence interval over 100 iterations. The models of interest are case-base with logistic regression (CBLR), Case-Base Neural Networks (CBNN), Cox, DeepHit, DeepSurv, Deep Survival Machines (DSM), Optimal (a CBLR model with the exact interaction terms and baseline hazard specified) and Kaplan-Meier (to serve as a baseline, predicting the average for all individuals)."}
#readRDS(here::here("figures", "megaPlot.rds"))
knitr::include_graphics(here::here("analyses/figures", "megaPlot.pdf"))
```

@@

METABRIC is a 30-yearlong study aiming to discover the molecular drivers of breast tumors, following 2000 individuals with breast cancer until death [@curtis2012genomic]. They described these growths as primary invasive breast carcinomas, with the goal of discovering both genetic and clinical risk factors for breast cancer survival [@curtis2012genomic]. We used the processed dataset made available through DeepSurv [@katzman2018DeepSurv], which includes 1980 patients of which 57.72% die due to breast cancer within a median 10 years of follow-up [@katzman2018DeepSurv]. There are 11 covariates in total: 4 RNA-Seq gene expressions (MKI67, EGFR, PGR and ERBB2) and 5 clinical features (hormone treatment indicator, radiotherapy indicator, chemotherapy indicator, ER-positive indicator and age at diagnosis) [@katzman2018DeepSurv].

@@

@@

This work was supported by subcontracts from UM1DK078616 and R01HL151855 to RS. We would like to thank Dr. James Meigs, Project Leader of these awards, for his support and helpful discussions. The work was also supported as part of the Congressionally Directed Medical Research Programs (CDMRP) award W81XWH-17-1-0347. We would also like to Thank Dr. James Hanley for his support and discussions while extending the case-base methodology.

>>>>

Now, let $b$ be the (user-defined) size of the *base series* and let $B$ be the sum of all follow-up times for the individuals in the study. If we sample the *base series* uniformly across the study base, then the hazard function of the sampling process is equal to $b/B$. Therefore, we have the following equality \footnote{We are abusing notation here, conflating hazards with probabilities. For a rigorous treatment, see Saarela \& Hanley (2015) section 3 \cite{saarela2015}.}: 
\begin{align}\label{eqn:main}
\frac{P\left(Y_i=1 \mid X_i, T_i\right)}{P\left(Y_i = 0 \mid X_i, T_i\right)} = \frac{h\left(T_i \mid X_i\right)}{b/B}.
\end{align}

To provide some intuition for this equation, the hazard is the rate at which the event ($Y_{i}=1$) occurs at a given time, conditional on your survival up to the follow-up-time of interest. As we are looking at person-moments, we may retrieve the rate at which an event does not occur ($Y_{i}=0)$ at a given time by taking the number of samples in $b$ and dividing it by the total number of moments in the study base $B$ at the current follow-up-time. Though an intuitive approach is used to describe this equation, a complete derivation of this equation can be found at [@saarela2015].

The odds of a person-moment being a part of the *case series* is the ratio of the hazard $h(T_i \mid X_i)$ and the uniform rate $b/B$. Sampling techniques are known to result in a loss of information. To address this concern, the original case-base methodology paper for logistic regression uses a measure for relative information when comparing two averages, based on the size of our base-series and case-series. If we let the size of $b=100c$, then the variances and covariances are expected to be proportional to $\frac{1}{c}+\frac{1}{100c}$ rather than $\frac{1}{c}+\frac{1}{\inf}$, about 1 percent larger. Concerns were raised about dependency due to multiple person-moments coming from each individual, the information loss was studied in Saarela 2015, where they found a ratio of 100:1, b:c, give an efficiency comparable to Mantel-Haenszel and conditional logistic regression methods (Saarela 2015). Using \eqref{eqn:main}, we can see how the log-hazard function can be estimated from the log-odds arising from case-base sampling: \begin{align}\label{eqn:offset}
\log \left( h\left(t \mid X_i\right)\right) = \log \left(\frac{P\left(Y_i = 1 \mid X_i, t\right)}{P\left(Y_i = 0 \mid X_i, t\right)}\right) + \log\left(\frac{b}{B}\right).
\end{align}

Sampling techniques are known to result in a loss of information. To address this concern, the original case-base methodology paper uses a measure for relative information when comparing two averages, based on the size of our base-series and case-series. If we let the size of $b=100c$, then the variances and covariances are expected to be proportional to $\frac{1}{c}+\frac{1}{100c}$ rather than $\frac{1}{c}+\frac{1}{\inf}$, about 1 percent larger. If there is a concern with a 1 percent inflation of these estimates, a larger ratio than 100 may be used. As there are concerns about dependency due multiple person-moments coming from each individual, the information loss was studied in Saarela 2015 ([@saarela2015]), where they found a ratio of 100:1 base-series to case series gives an efficiency comparable to Mantel-Haenszel and conditional logistic regression methods [@saarela2015].

To estimate the correct hazard function, we adjusting for the bias introduced when sampling a fraction of the study base $B$ by adding an offset term $\log\left(\frac{B}{b} \right)$ \eqref{eqn:offset}. Next, we propose using neural networks to model the odds.

@@

After case-base sampling, we pass all features, including time, into any user-defined feed-forward component, to which an offset term is added, then passed through a sigmoid activation function (Figure \ref{fig:NNarch}). We use the sigmoid activation function as it is the inverse of the log-odds and commonly available in popular neural network packages (tensorflow, keras, pytorch, etc) [REF?]. which we can use to calculate the hazard. The general form for the neural network using CBNN is:

@@

where $T$ is a random variable representing the event time, $X$ is the random variable for a covariate profile, $f_{\theta}(X, T)$ represents any feed-forward neural network architecture, $\log\left(\frac{B}{b}\right)$ is bias term set by case-base sampling, $\theta$ is the set of parameters learned by the neural network and $\sigmoid(x)=\frac{1}{1+e^{-x}}$. By approximating a higher-order polynomial of time using a neural network, the baseline hazard specification is now data-driven, where user-defined hyperparameters such as regularization, number of layers and nodes control the flexibility of the hazard function. We provide a detailed description of the choices we made in the next sub-section.

@@

We use two metrics to assess the performance of the different methods of interest: 1) (IPA) [@kattan2018index] and 2) inverse probability censoring weights-adjusted concordance index ($C_{IPCW}$) [@uno2011], which we define below. We may be interested in an absolute risk a specific number of years in the future. For the sake of comparison, we plot each metric over time $t$ so that the performance is transparent. As we go further along follow-up-time, we expect the performance of all models to drop as there are fewer individuals left in the study by then. For $C_{IPCW}$, the score demonstrates performance up to time $t$  [@uno2011], while IPA is performance at time $t$.

@@

```{r megaPlot, echo = FALSE , fig.align = 'center', out.width = "100%", fig.cap = "Summarizes the simple simulation (A, B), complex simulation (C, D), SUPPORT case study (E, F) and METABRIC case study (G, H) results. The first row shows the IPA score for each model in each study over follow-up time. Negative values mean our model performs worse than the null model and positive values mean the model performs better. The second row demonstrates the $C_{IPCW}$ score for each model in each study over follow-up time. A score of 1 is the maximum performance for either metric. Each model-specific metric in each study shows a 95-percent confidence interval over 100 iterations. The models of interest are case-base with logistic regression (CBLR), Case-Base Neural Networks (CBNN), Cox, DeepHit, DeepSurv, Deep Survival Machines (DSM), Optimal (a CBLR model with the exact interaction terms and baseline hazard specified) and Kaplan-Meier (to serve as a baseline, predicting the average for all individuals). Note than Cox and CBLR are superimposed."}
#readRDS(here::here("figures", "megaPlot.rds"))
knitr::include_graphics(here::here("analyses/figures", "megaPlot.pdf"))
```

@@

METABRIC is a 30-yearlong study aiming to discover the molecular drivers of breast tumors, following 2000 individuals with breast cancer until death [@curtis2012genomic]. They described these growths as primary invasive breast carcinomas, with the goal of discovering both genetic and clinical risk factors for breast cancer survival [@curtis2012genomic]. We used the processed dataset made available through DeepSurv [@katzman2018DeepSurv], which includes 1980 patients of which 57.72% die due to breast cancer within a median 10 years of follow-up [@katzman2018DeepSurv]. There are 10 covariates in total: time, 4 RNA-Seq gene expressions (MKI67, EGFR, PGR and ERBB2) and 5 clinical features (hormone treatment indicator, radiotherapy indicator, chemotherapy indicator, ER-positive indicator and age at diagnosis) [@katzman2018DeepSurv].

@@

The way neural networks are used in this paper are not directly interpretable. This is by design, as we only wish to compare to competitors based on predictive ability. All feed-forward neural network models can make use of different architectures to improve performance or interpretability. While a deep neural network as used in this paper is not directly interpretable, techniques such as Local Interpretable Model-Agnostic Explanations (LIME) [@ribeiro2016model] may be used to interpret the relevance of each feature to a given prediction. An example of architectural interpretability is a convolutional neural network for images, which provide positional information about the relevant pixels in an image [@keras]. Note that neither of these are specific to CBNN. As CBNN is concerned with learning time-varying interactions, a concern may be interpreting which time-varying interactions have been learned. Other secondary methods after the fitting process may be used to retrieve said information https://ui.adsabs.harvard.edu/abs/2017arXiv170504977T/abstract.

@@

This work was supported by subcontracts from UM1DK078616 and R01HL151855 to Dr. Rob Sladek. We would like to thank Dr. James Meigs, the project leader of these awards, for his support and helpful discussions. The work was also supported as part of the Congressionally Directed Medical Research Programs (CDMRP) award W81XWH-17-1-0347. We would also like to thank Dr. James Hanley for his support and discussions while extending the case-base methodology.

### 2022-11-15 SIMPaper/SIMPaper.tex
addedStatTemplate

where \(T\) is a random variable representing the event time, \(X\) is
the random variable for a covariate profile, \(f_{\theta}(X, T)\)
represents any feed-forward neural network architecture,
\(\log\left(\frac{B}{b}\right)\) is bias term set by case-base sampling,
\(\theta\) is the set of parameters learned by the neural network and
\(\mathrm{sigmoid}(x)=\frac{1}{1+e^{-x}}\). By approximating a
higher-order polynomial of time using a neural network, the baseline
hazard specification is now data-driven, where user-defined
hyperparameters such as regularization, number of layers and nodes
control the flexibility of the fukk hazard function. We provide a
detailed description of the choices we made in the next sub-section.

@@

METABRIC is a 30-yearlong study aiming to discover the molecular drivers
of breast tumors, following 2000 individuals with breast cancer until
death \citep{curtis2012genomic}. They described these growths as primary
invasive breast carcinomas, with the goal of discovering both genetic
and clinical risk factors for breast cancer survival
\citep{curtis2012genomic}. We used the processed dataset made available
through DeepSurv \citep{katzman2018DeepSurv}, which includes 1980
patients of which 57.72\% die due to breast cancer within a median 10
years of follow-up \citep{katzman2018DeepSurv}. There are 11 covariates
in total: 4 RNA-Seq gene expressions (MKI67, EGFR, PGR and ERBB2) and 5
clinical features (hormone treatment indicator, radiotherapy indicator,
chemotherapy indicator, ER-positive indicator and age at diagnosis)
\citep{katzman2018DeepSurv}.

@@

This work was supported by subcontracts from UM1DK078616 and R01HL151855
to RS. We would like to thank Dr.~James Meigs, Project Leader of these
awards, for his support and helpful discussions. The work was also
supported as part of the Congressionally Directed Medical Research
Programs (CDMRP) award W81XWH-17-1-0347. We would also like to Thank
Dr.~James Hanley for his support and discussions while extending the
case-base methodology.

>>>>

where \(T\) is a random variable representing the event time, \(X\) is
the random variable for a covariate profile, \(f_{\theta}(X, T)\)
represents any feed-forward neural network architecture,
\(\log\left(\frac{B}{b}\right)\) is bias term set by case-base sampling,
\(\theta\) is the set of parameters learned by the neural network and
\(\mathrm{sigmoid}(x)=\frac{1}{1+e^{-x}}\). By approximating a
higher-order polynomial of time using a neural network, the baseline
hazard specification is now data-driven, where user-defined
hyperparameters such as regularization, number of layers and nodes
control the flexibility of the hazard function. We provide a detailed
description of the choices we made in the next sub-section.

@@

METABRIC is a 30-yearlong study aiming to discover the molecular drivers
of breast tumors, following 2000 individuals with breast cancer until
death \citep{curtis2012genomic}. They described these growths as primary
invasive breast carcinomas, with the goal of discovering both genetic
and clinical risk factors for breast cancer survival
\citep{curtis2012genomic}. We used the processed dataset made available
through DeepSurv \citep{katzman2018DeepSurv}, which includes 1980
patients of which 57.72\% die due to breast cancer within a median 10
years of follow-up \citep{katzman2018DeepSurv}. There are 10 covariates
in total: time, 4 RNA-Seq gene expressions (MKI67, EGFR, PGR and ERBB2)
and 5 clinical features (hormone treatment indicator, radiotherapy
indicator, chemotherapy indicator, ER-positive indicator and age at
diagnosis) \citep{katzman2018DeepSurv}.

@@

This work was supported by subcontracts from UM1DK078616 and R01HL151855
to Dr.~Rob Sladek. We would like to thank Dr.~James Meigs, the project
leader of these awards, for his support and helpful discussions. The
work was also supported as part of the Congressionally Directed Medical
Research Programs (CDMRP) award W81XWH-17-1-0347. We would also like to
thank Dr.~James Hanley for his support and discussions while extending
the case-base methodology.
